Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-13 experiments/notebooks/poc.qmd
poc chart done

temp = 0.3
p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
plt = display(Plots.plot(p0, p1, size=(800,320), layout=(1,2)))
```

@@

λ₁ = 0.2
λ₂ = 0.5
λ₃ = 3.0
Λ = [λ₁, λ₂, λ₃]
η = 0.01

>>>>

temp = 0.1
p0 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_size=true, zoom=0, temp=temp)
p1 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_set_loss=true, zoom=0, temp=temp)
p2 = Plots.contourf(mach.model, mach.fitresult, permutedims(X), labels; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
plt = display(Plots.plot(p0, p1, p2, size=(1400,320), layout=(1,3)))
# savefig(joinpath(output_images_path, "poc_set_size.png"))
```

@@

λ₁ = 0.2
λ₂ = 1.5
λ₃ = 3.0
Λ = [λ₁, λ₂, λ₃]
η = 0.01

### 2023-12-13 paper/camera-ready/altmeyer.tex
poc chart done

Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data. While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations. The technical appendix provides additional implementation details for any tasks related to energy-based modelling.

@@

The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself. This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$. Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the technical appendix.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other benchmark datasets (Tables~\ref{tab:results-linearly-separable} to~\ref{tab:results-fashion-mnist} in the technical appendix).

@@

Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance. Upon visual inspection of the generated images we actually find that \textit{ECCCo+} performs much better than \textit{REVISE} (see appendix). Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data. For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline. This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).

>>>>

Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data. While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations. The appendix provides additional implementation details for any tasks related to energy-based modeling.

@@

The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself. This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$. Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the appendix.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other benchmark datasets (additional results can be found in the appendix).

@@

Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance. Upon visual inspection of the generated images we actually find that \textit{ECCCo+} performs much better than \textit{REVISE} (see appendix). Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data. For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline. This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness.
